Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-01-14 Advice Giving.Rmd
Added linebreaks for easier change tracking

# Introduction
Seeking and giving advice can be challenging conversations to navigate. Advice seekers may be embarrassed to ask for help [@Benabou2022], and they may fail to do so if the information they receive could lead to disappointment [@Gill2012]. Indeed, people prefer advisers who withhold unpleasant information [@Shalvi2019] and punish those who tell them things they do not want to hear [@John2019]. People appear to engage in strategic selection of advice to avoid unpleasant information and protect the utility they get from positive beliefs about themselves and the future they anticipate [@Loewenstein2018; @Loewenstein1987]. But do advice givers anticipate and consider the hedonic cost of honest information to the recipient? While previous work has emphasized the instrumental value of advice (e.g., satisfaction with a decision or accuracy of a prediction), we focus on the motivations of the person providing advice.

@@

# Open Science Statement
We report all sample sizes, data exclusions, all manipulations, and all measures in the studies. Screen captures of the experimental materials are available in the Supplemental Information. The complete data, code to reproduce all statistical analyses and figures in the manuscript, as well as the preregistration reports are available via OSF. All our studies were preregistered on AsPredicted.org. (will do this later)

# Study 1
We begin by examining whether advisers take the expectations of advisees into account, even when those expectations are not relevant to the quality of advice. In a three-stage experiment, we first recruit a sample of advisees and ask them to complete a mathematics quiz. After answering all ten multiple-choice questions, participants are anchored to low or high performance expectations and asked to guess how many questions they answered correctly. Next, we recruit advice givers and show them either only the true performance of the advisee, or the true performance and how many questions they guessed they answered correctly. Advisers are tasked with recommending whether the advisee should compete against a group of high-performers or a group of low-performers on the mathematics quiz. Finally, we invite the advisees back and show them the advice they have received. Advisees then pick whether to compete against high or low performers, based on their past performance on the quiz. Our key hypothesis is that advisees who were anchored to high expectations would be more likely to receive advice to compete against the high performer group.

## Methods
```{r}

@@

Next, we recruited `r scales::comma(dim(study1_time1All)[1])` participants in Stage 1 of our main experiment, for the role of advisees. To arrive at a gender-balanced sample, we dropped the last two male participants to complete the survey, ending up with a sample of 500 men and 500 women ($M_{\text{Age}}$ = `r mean(study1_time1$Age)`,50\% Female). Participants completed the same 10-item mathematics quiz as the earlier participants and were informed that their performance would affect their bonus earnings in a follow-up study to be conducted a few days later. After completing the quiz, we informed them of the average score of a group of five participants from the preliminary survey. We randomly assigned them to learn about the 5 th percentile of groups, which scored `r p5`. (“Low Expectations” treatment) or the 95 th percentile of groups, which scored `r p95` (“High Expectations” treatment). Participants then made a guess (unincentivized) about how many questions they think they answered correctly. The survey concluded with basic demographic questions.

@@

## Results
We begin by examining the performance of the Stage 1 participants. Because our treatment takes place after participants complete the mathematics quiz, we would not expect a difference in performance across the Low and High Expectations treatments. Indeed, the two groups scored no different from one another (`r round(mean(study1_time1[study1_time1$Condition == "Low",]$count),2)` and `r round(mean(study1_time1[study1_time1$Condition == "High",]$count),2)` for Low and High Expectations treatments, respectively, `r papaja::apa_print(t.test(count ~ Condition, data = study1_time1, var.equal = T))$statistic`). The expectations treatment did, however, affect how well they thought they performed. Participants in the “Low Expectations” treatment guessed a score of `r round(mean(study1_time1[study1_time1$Condition == "Low",]$count),2)` vs. `r round(mean(study1_time1[study1_time1$Condition == "High",]$Guess),2)` in the High Expectations treatment (`r papaja::apa_print(t.test(Guess ~ Condition, data = study1_time1, var.equal = T))$statistic`), showing that the manipulation was successful. Contrary to our expectations, we did find a gender difference in performance: men scored `r round(mean(study1_time1[study1_time1$Gender == "Male",]$count),2)` on average, while women scored `r round(mean(study1_time1[study1_time1$Gender == "Female",]$count),2)` (`r papaja::apa_print(t.test(count ~ Gender, data = study1_time1, var.equal = T))$statistic`). This difference, however, does not affect the interpretation of our findings, which will rely on an interaction of gender with an experimental treatment for Stage 2 participants. Consistent with this difference, men thought they answered more questions correctly than did women (`r round(mean(study1_time1[study1_time1$Gender == "Male",]$Guess),2)` vs `r round(mean(study1_time1[study1_time1$Gender == "Female",]$Guess),2)`, `r papaja::apa_print(t.test(Guess ~ Gender, data = study1_time1, var.equal = T))$statistic`).

@@

```
Next, we examine whether advisers took the advisee’s expectations into account. Column 1 of Table \@ref(tab:study1regs) reports a linear probability model on advising to compete against the High Performer group for advisers in the “Expectations” treatment. Because each adviser was paired with ten advisees, we cluster standard errors at the adviser level. Consistent with our prediction, we find that advisees in the High Expectation condition are more likely to be recommended to compete against the High Performer Group, but the result is not significant.

To test whether expectations can lead to more flattering advice for men than for women, we report a linear probability model with advice to compete against the high group as the outcome measure, and the advisee’s gender, whether expectations were shown to the adviser, and the interaction of the two (Column 2 of Table \@ref(tab:study1regs)). As predicted, we find a significant interaction effect: men are more likely to be advised to compete against the High Performers when expectations are shown than when they are hidden (p < 0.001; see Figure \@ref(fig:study1genderdiff),).

To examine the quality of advice, we computed the expected bonus earnings for all recommendations. We made this calculation by assuming that participants adopted the advice and competed against the recommended competitor group. We then compared their scores with all 20 possible competitors one by one and calculated the bonus accordingly to determine the expected bonus earnings. We conducted an OLS regression with whether givers see the expectation (Baseline or Information condition) and the gender of the Advisee (Female or Male), as well as their interaction. Supporting our theory, male participants will receive less bonus given advice they receive (b = -0.005, p < .01; see also Column 3 of Table \@ref(tab:study1regs)

@@

```
Finally, we look at the actual earnings of advisees. Although participants in the High Expectation condition earned less, this difference is not significant (`r model2_apa$estimate['AdviseeConditionHigh']`, `r model2_apa$statistic['AdviseeConditionHigh']`).

@@

## Discussion
In Study 1, we discovered that although advisers are encouraged to provide accurate advice, they still take into account the expectations of advisees, which are irrelevant to the utility of the advice. Due to the fact that males tend to be more confident than their equally competent female counterparts, they are more likely to be recommended to compete against high-performance competitors. However, when advice is given to match the advisees' expectations, these flattering recommendations result in lower expected bonus earnings for male participants.

# Study 2
We then conduct a three-stage experiment to determine why advisers give flattering advice. First, we ask a group of advisees to upload selfies and predict their rank in a group of ten people of the same gender based on physical attractiveness. Next, we recruit and ask advisers to rank selfies of the opposite gender and recommend which rank the participant ranked 7th should bet on, with the bonus being contingent on whether the participant guessed accurately or found the adviser likable. Finally, we show the advisees the advice they have received and ask them to make their final prediction about their rank and evaluate the advice giver in terms of likability, warmth, and trustworthiness. Our main hypothesis is that advisers who want to be liked by advisees would recommend a higher rank.

@@

## Methods
We recruited 300 participants and invited them to upload photos of themselves (selfies) to be rated by other participants on attractiveness. We ended up with a sample of `r scales::comma(dim(study2_time1[study2_time1$gender == 'Male',])[1])` men and `r scales::comma(dim(study2_time1[study2_time1$gender == 'Female',])[1])` women who agreed to do so and uploaded usable pictures. We selected the first 100 photos from women to arrive at a gender-balanced sample ($M_{\text{Age}}$ = `r mean(study2_time1_200$age)` years; 50.0\% Female). We informed them that their selfies will be randomly grouped into nine others from the same gender and ranked by people of the opposite gender and asked them to predict their rank, with 1st rank represents the most attractive and 10th rank represents the least attractive. We assigned each participant to a group of 10 participants of their own gender.

Next, we recruited `r scales::comma(dim(study2_time2)[1])`new participants and asked them to rank the attractiveness of one such group of 10 participants of the opposite gender ($M_{\text{Age}}$ = `r mean(study2_time2$age)` years; `r round(mean(study2_time2$gender == "Female")*100, 2)`\% Female). After they did so, we showed them the photo of the participant they ranked as the 7th  most attractive (that is, the 4th  least attractive) and reminded them that they had ranked this person at 7th attractive. We (truthfully) informed them that this participant would have a chance to earn a bonus if they accurately guessed their rank based on the aggregated ratings of all participants who had ranked this group. Participants were then randomly assigned to an “accuracy” or a “likeability” treatment. In the accuracy treatment, they were told that they would receive a bonus ($1) identical to that of the advisee. While in the “likeability” treatment, the amount their bonus depended on how the advisee rated them in terms of likeability, after observing only the rank that they had advised the advisee to guess. Advisees would rate to what extent they think the adviser is likable on a 5-point scale, and each point they earn increases their bonus by 20 cents.

Finally, we invited participants from Stage 1 and were ranked as the 7th attractive by at least one adviser (so that they received advice) back for the follow-up survey. Following our preregistration, we kept the survey open for 7 days. In total, `r scales::comma(dim(study2_time3)[1])`  participants (`r scales::comma(dim(study2_time3[study2_time3$gender == 'Male',])[1])` men, `r scales::comma(dim(study2_time3[study2_time3$gender == 'Female',])[1])` women) returned. They were reminded of the selfie they uploaded in Stage 1 and informed that a group of 10 selfies, including theirs, had been rated by other participants from Prolific. They were also introduced to the opportunity to earn a $1 bonus if they guessed their rank accurately. Participants were then shown the advice from a randomly selected adviser and made their predictions. After deciding which rank to bet on, advisees were reminded of the advice they received and asked to rate the advice giver's likability, warmth, friendliness, good-naturedness, trustworthiness, and sincerity on a 5-point scale. The last 6 items were adopted from @Fiske2007.

## Results
```{r study2advice, fig.cap="Advice for 7th most attractive participants"}
library(dplyr)
library(ggplot2)

@@

To test whether flattering advice brings more positive evaluation of advisers, we create a single index for warmth and trustworthiness by taking average of ratings on warmth, friendliness, good-naturedness, trustworthiness for the former and ratings on trustworthiness, and sincerity for the later. As seen in Figure \@ref(fig:study2rating), we found that advisees who suggested that the advisee was more attractive (lower rank) were indeed rated as more likeable and warm (b = -0.109, p < .05; b = -0.133, p < .001; See Columns 1 and 2 of Table 2). Interestingly, these benefits are not at the cost of being hypocritical; advisers are also seen as more trustworthy, but the result is only directional (Column 3 of Table 2). Notably, however, we were not powered to do a comparison across the two experimental groups.

# Discussion
dvice substantially shapes people's career and personal outcomes. Therefore, understanding advisers' motivations has far-reaching implications. We propose that advisers' focus on advisees' expectations and the desire to avoid disappointing them can induce gender differences in the advice given and ultimately in the outcomes achieved. Gender differences in career outcomes contribute significantly to the gender wage gap, with women often holding less senior positions and experiencing worse career outcomes, even with equal performance. Therefore, identifying a factor such as expectations that contributes to this gap has important implications for organizational practice.

>>>>

# Introduction

Seeking and giving advice can be challenging conversations to navigate.
Advice seekers may be embarrassed to ask for help [@Benabou2022], and
they may fail to do so if the information they receive could lead to
disappointment [@Gill2012]. Indeed, people prefer advisers who withhold
unpleasant information [@Shalvi2019] and punish those who tell them
things they do not want to hear [@John2019]. People appear to engage in
strategic selection of advice to avoid unpleasant information and
protect the utility they get from positive beliefs about themselves and
the future they anticipate [@Loewenstein2018; @Loewenstein1987]. But do
advice givers anticipate and consider the hedonic cost of honest
information to the recipient? While previous work has emphasized the
instrumental value of advice (e.g., satisfaction with a decision or
accuracy of a prediction), we focus on the motivations of the person
providing advice.

@@

# Open Science Statement

We report all sample sizes, data exclusions, all manipulations, and all
measures in the studies. Screen captures of the experimental materials
are available in the Supplemental Information. The complete data, code
to reproduce all statistical analyses and figures in the manuscript, as
well as the preregistration reports are available via OSF. All our
studies were preregistered on AsPredicted.org. (will do this later)

# Study 1

We begin by examining whether advisers take the expectations of advisees
into account, even when those expectations are not relevant to the
quality of advice. In a three-stage experiment, we first recruit a
sample of advisees and ask them to complete a mathematics quiz. After
answering all ten multiple-choice questions, participants are anchored
to low or high performance expectations and asked to guess how many
questions they answered correctly. Next, we recruit advice givers and
show them either only the true performance of the advisee, or the true
performance and how many questions they guessed they answered correctly.
Advisers are tasked with recommending whether the advisee should compete
against a group of high-performers or a group of low-performers on the
mathematics quiz. Finally, we invite the advisees back and show them the
advice they have received. Advisees then pick whether to compete against
high or low performers, based on their past performance on the quiz. Our
key hypothesis is that advisees who were anchored to high expectations
would be more likely to receive advice to compete against the high
performer group.

## Methods

@@

Next, we recruited `r scales::comma(dim(study1_time1All)[1])`
participants in Stage 1 of our main experiment, for the role of
advisees. To arrive at a gender-balanced sample, we dropped the last two
male participants to complete the survey, ending up with a sample of 500
men and 500 women ($M_{\text{Age}}$ = `r mean(study1_time1$Age)`,50%
Female). Participants completed the same 10-item mathematics quiz as the
earlier participants and were informed that their performance would
affect their bonus earnings in a follow-up study to be conducted a few
days later. After completing the quiz, we informed them of the average
score of a group of five participants from the preliminary survey. We
randomly assigned them to learn about the 5 th percentile of groups,
which scored `r p5`. (“Low Expectations” treatment) or the 95 th
percentile of groups, which scored `r p95` (“High Expectations”
treatment). Participants then made a guess (unincentivized) about how
many questions they think they answered correctly. The survey concluded
with basic demographic questions.

@@

## Results

We begin by examining the performance of the Stage 1 participants.
Because our treatment takes place after participants complete the
mathematics quiz, we would not expect a difference in performance across
the Low and High Expectations treatments. Indeed, the two groups scored
no different from one another
(`r round(mean(study1_time1[study1_time1$Condition == "Low",]$count),2)`
and
`r round(mean(study1_time1[study1_time1$Condition == "High",]$count),2)`
for Low and High Expectations treatments, respectively,
`r papaja::apa_print(t.test(count ~ Condition, data = study1_time1, var.equal = T))$statistic`).
The expectations treatment did, however, affect how well they thought
they performed. Participants in the “Low Expectations” treatment guessed
a score of
`r round(mean(study1_time1[study1_time1$Condition == "Low",]$count),2)`
vs.
`r round(mean(study1_time1[study1_time1$Condition == "High",]$Guess),2)`
in the High Expectations treatment
(`r papaja::apa_print(t.test(Guess ~ Condition, data = study1_time1, var.equal = T))$statistic`),
showing that the manipulation was successful. Contrary to our
expectations, we did find a gender difference in performance: men scored
`r round(mean(study1_time1[study1_time1$Gender == "Male",]$count),2)` on
average, while women scored
`r round(mean(study1_time1[study1_time1$Gender == "Female",]$count),2)`
(`r papaja::apa_print(t.test(count ~ Gender, data = study1_time1, var.equal = T))$statistic`).
This difference, however, does not affect the interpretation of our
findings, which will rely on an interaction of gender with an
experimental treatment for Stage 2 participants. Consistent with this
difference, men thought they answered more questions correctly than did
women
(`r round(mean(study1_time1[study1_time1$Gender == "Male",]$Guess),2)`
vs
`r round(mean(study1_time1[study1_time1$Gender == "Female",]$Guess),2)`,
`r papaja::apa_print(t.test(Guess ~ Gender, data = study1_time1, var.equal = T))$statistic`).

@@

Next, we examine whether advisers took the advisee’s expectations into
account. Column 1 of Table \@ref(tab:study1regs) reports a linear
probability model on advising to compete against the High Performer
group for advisers in the “Expectations” treatment. Because each adviser
was paired with ten advisees, we cluster standard errors at the adviser
level. Consistent with our prediction, we find that advisees in the High
Expectation condition are more likely to be recommended to compete
against the High Performer Group, but the result is not significant.

To test whether expectations can lead to more flattering advice for men
than for women, we report a linear probability model with advice to
compete against the high group as the outcome measure, and the advisee’s
gender, whether expectations were shown to the adviser, and the
interaction of the two (Column 2 of Table \@ref(tab:study1regs)). As
predicted, we find a significant interaction effect: men are more likely
to be advised to compete against the High Performers when expectations
are shown than when they are hidden (p \< 0.001; see Figure
\@ref(fig:study1genderdiff),).

To examine the quality of advice, we computed the expected bonus
earnings for all recommendations. We made this calculation by assuming
that participants adopted the advice and competed against the
recommended competitor group. We then compared their scores with all 20
possible competitors one by one and calculated the bonus accordingly to
determine the expected bonus earnings. We conducted an OLS regression
with whether givers see the expectation (Baseline or Information
condition) and the gender of the Advisee (Female or Male), as well as
their interaction. Supporting our theory, male participants will receive
less bonus given advice they receive (b = -0.005, p \< .01; see also
Column 3 of Table \@ref(tab:study1regs)

@@

Finally, we look at the actual earnings of advisees. Although
participants in the High Expectation condition earned less, this
difference is not significant
(`r model2_apa$estimate['AdviseeConditionHigh']`,
`r model2_apa$statistic['AdviseeConditionHigh']`).

@@

## Discussion

In Study 1, we discovered that although advisers are encouraged to
provide accurate advice, they still take into account the expectations
of advisees, which are irrelevant to the utility of the advice. Due to
the fact that males tend to be more confident than their equally
competent female counterparts, they are more likely to be recommended to
compete against high-performance competitors. However, when advice is
given to match the advisees' expectations, these flattering
recommendations result in lower expected bonus earnings for male
participants.

# Study 2

We then conduct a three-stage experiment to determine why advisers give
flattering advice. First, we ask a group of advisees to upload selfies
and predict their rank in a group of ten people of the same gender based
on physical attractiveness. Next, we recruit and ask advisers to rank
selfies of the opposite gender and recommend which rank the participant
ranked 7th should bet on, with the bonus being contingent on whether the
participant guessed accurately or found the adviser likable. Finally, we
show the advisees the advice they have received and ask them to make
their final prediction about their rank and evaluate the advice giver in
terms of likability, warmth, and trustworthiness. Our main hypothesis is
that advisers who want to be liked by advisees would recommend a higher
rank.

@@

## Methods

We recruited 300 participants and invited them to upload photos of
themselves (selfies) to be rated by other participants on
attractiveness. We ended up with a sample of
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Male',])[1])`
men and
`r scales::comma(dim(study2_time1[study2_time1$gender == 'Female',])[1])`
women who agreed to do so and uploaded usable pictures. We selected the
first 100 photos from women to arrive at a gender-balanced sample
($M_{\text{Age}}$ = `r mean(study2_time1_200$age)` years; 50.0% Female).
We informed them that their selfies will be randomly grouped into nine
others from the same gender and ranked by people of the opposite gender
and asked them to predict their rank, with 1st rank represents the most
attractive and 10th rank represents the least attractive. We assigned
each participant to a group of 10 participants of their own gender.

Next, we recruited `r scales::comma(dim(study2_time2)[1])`new
participants and asked them to rank the attractiveness of one such group
of 10 participants of the opposite gender ($M_{\text{Age}}$ =
`r mean(study2_time2$age)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female). After
they did so, we showed them the photo of the participant they ranked as
the 7th most attractive (that is, the 4th least attractive) and reminded
them that they had ranked this person at 7th attractive. We (truthfully)
informed them that this participant would have a chance to earn a bonus
if they accurately guessed their rank based on the aggregated ratings of
all participants who had ranked this group. Participants were then
randomly assigned to an “accuracy” or a “likeability” treatment. In the
accuracy treatment, they were told that they would receive a bonus (\$1)
identical to that of the advisee. While in the “likeability” treatment,
the amount their bonus depended on how the advisee rated them in terms
of likeability, after observing only the rank that they had advised the
advisee to guess. Advisees would rate to what extent they think the
adviser is likable on a 5-point scale, and each point they earn
increases their bonus by 20 cents.

Finally, we invited participants from Stage 1 and were ranked as the 7th
attractive by at least one adviser (so that they received advice) back
for the follow-up survey. Following our preregistration, we kept the
survey open for 7 days. In total,
`r scales::comma(dim(study2_time3)[1])` participants
(`r scales::comma(dim(study2_time3[study2_time3$gender == 'Male',])[1])`
men,
`r scales::comma(dim(study2_time3[study2_time3$gender == 'Female',])[1])`
women) returned. They were reminded of the selfie they uploaded in Stage
1 and informed that a group of 10 selfies, including theirs, had been
rated by other participants from Prolific. They were also introduced to
the opportunity to earn a \$1 bonus if they guessed their rank
accurately. Participants were then shown the advice from a randomly
selected adviser and made their predictions. After deciding which rank
to bet on, advisees were reminded of the advice they received and asked
to rate the advice giver's likability, warmth, friendliness,
good-naturedness, trustworthiness, and sincerity on a 5-point scale. The
last 6 items were adopted from @Fiske2007.

## Results

```{r study2advice, fig.cap="Advice for 7th most attractive participants"}
library(dplyr)
library(ggplot2)

@@

To test whether flattering advice brings more positive evaluation of
advisers, we create a single index for warmth and trustworthiness by
taking average of ratings on warmth, friendliness, good-naturedness,
trustworthiness for the former and ratings on trustworthiness, and
sincerity for the later. As seen in Figure \@ref(fig:study2rating), we
found that advisees who suggested that the advisee was more attractive
(lower rank) were indeed rated as more likeable and warm (b = -0.109, p
\< .05; b = -0.133, p \< .001; See Columns 1 and 2 of Table 2).
Interestingly, these benefits are not at the cost of being hypocritical;
advisers are also seen as more trustworthy, but the result is only
directional (Column 3 of Table 2). Notably, however, we were not powered
to do a comparison across the two experimental groups.

# Discussion

dvice substantially shapes people's career and personal outcomes.
Therefore, understanding advisers' motivations has far-reaching
implications. We propose that advisers' focus on advisees' expectations
and the desire to avoid disappointing them can induce gender differences
in the advice given and ultimately in the outcomes achieved. Gender
differences in career outcomes contribute significantly to the gender
wage gap, with women often holding less senior positions and
experiencing worse career outcomes, even with equal performance.
Therefore, identifying a factor such as expectations that contributes to
this gap has important implications for organizational practice.
